---
title: "EDA IEOR 142, Final Project"
date: "December 16, 2019"
output: pdf_document
---

```{r}
library(softImpute)
library(gridExtra, verbose=FALSE, warn.conflicts=FALSE, quietly=TRUE)
library(randomForest)
library(ranger)
library(dplyr)
library(reshape2)
library(caTools)
library(ROCR)
library(MASS)
library(ggplot2)
```

```{r}
restore = list(repr.plot.width=8, repr.plot.height=3)
PALETTE = c("#00A9FF", "#F8766D", "#7CAE00", "#C77CFF", "#CD9600", "#00BE67", "#FF61CC", "#00BFC4")
theme.x_axis_only = theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.grid.major.y=element_blank(), panel.grid.minor.y=element_blank())
theme.no_legend = theme(legend.position="none")
theme.legend_title = theme(legend.title=element_text(size=7))
data <- read.csv("us_suicides_merged_no_na.csv")
```

#EDA

Our cleaned and merged data consists of 372 observations and 14 variables.
```{r}
nrow(data)
ncol(data)
```

```{r}
#First look at the first 6 and last 6 observations of our data
head(data)
tail(data)
```
The dataset contains 31 unique years from 1985 to 2015, the suicide rate per 100k has a variance of 175.0296.
```{r}
unique(sort(data$year))
length(unique(sort(data$year)))
var(data$suicides.100k.pop)
#table(data$suicides.100k.pop) / length(data$suicides.100k.pop)) # relative frequencies
as.numeric(names(table(data$suicides.100k.pop))[table(data$suicides.100k.pop) == max(table(data$suicides.100k.pop))]) # mode for suicide rate
as.numeric(names(table(data$gdp_per_capita....))[table(data$gdp_per_capita....) == max(table(data$gdp_per_capita....))]) # mode for gdp per capita
```

```{r}
range(data$suicides.100k.pop)
data[data$suicides.100k.pop == min(data$suicides.100k.pop), ]
data[data$suicides.100k.pop == max(data$suicides.100k.pop), ]
```
###Investigating Suicide rate and Sex
There are 186 males and 186 females.
There is also 62 records for every age range provided in the data. The data seems to be split evenly thus far except for the  generation variable.Generation X has the highest amount of records and Generation Z has the least. The Suicide rate had a decline from about the late 1990's to the mid 2000's but has been steadily increasing since around the year 2008.
```{r}
table(data$sex)
table(data$age) 
table(data$generation)
max(table(data$generation))
ggplot(data)  + ggtitle("suicide trend over years for the US") +
geom_col(aes(x=data$year, y=data$suicides.100k.pop)) + xlab("Year") + ylab("suicide number") 
```

```{r}
p1 = ggplot(data)  + ggtitle("suicide number per age for the US") +
geom_col(aes(x=data$sex, y=data$suicides.100k.pop)) + xlab("Sex") + ylab("suicide number") 


p2= ggplot(data, aes(x=data$year, y=data$suicides.100k.pop, fill=data$sex), xlab("Year"), ylab("Suicide Rate")) +
   geom_bar(stat="identity", width=1, position = "dodge")  
grid.arrange(p1, p2, nrow=2, ncol = 1.2)

ggplot(data) + geom_col(aes(x=year, y=suicides.100k.pop, width=0.7, fill=factor(sex, levels=c("male", "female")))) + ggtitle("Suicide Rate by Gender") + xlab("Year") + ylab("Suicide Rate") +  scale_fill_manual(values = c("#00A9FF", "#F8766D"), name = "Gender")

```


###Investigating Suicide rate and Age

Suicide rates are highest among indivduals in the age group 75+( This is mostly people considered to be from the G.I. generation(1901-1924) and Silent generation(1925-1945)) and the lowest rates occur in the age group 5-14(generation X and generation Z).
```{r}
p3 = ggplot(data) + ggtitle("Boxplot of Suicide Rate Per age range") + geom_boxplot(aes(x= data$age, y=data$suicides.100k.pop, fill = data$generation)) +
xlab("Age Range") + ylab("suicide rate")

p4 = ggplot(data) + ggtitle("Stacked, Suicide Rate Per age range") + geom_col(aes(x= data$age, y=data$suicides.100k.pop, fill = data$generation)) +
xlab("Age Range") + ylab("suicide rate")
grid.arrange(p3, p4, nrow=2, ncol = 1.1)
```

```{r}
ggplot(data)  + ggtitle("Suicide trend over generations, US") +
geom_col(aes(x= data$generation, y= data$suicides.100k.pop, fill=data$age), position="stack") + xlab("Generation") + ylab("Suicide Rate")
```


###Investigating Suicide rate and Population
```{r}
p5 = ggplot(data)  + ggtitle("Density Plot of population vs suicide rate") +
geom_density_2d(aes(x=data$population, y=data$suicides.100k.pop)) + xlab("population") + ylab("suicide rate")

p6 = ggplot(data)  + ggtitle("Area plot of population vs suicide rate") +
geom_area(aes(x=data$population, y=data$suicides.100k.pop)) + xlab("population") + ylab("suicide rate")

grid.arrange(p5, p6, nrow=2, ncol = 1.1)


```

``` {r}
ggplot(data) + ggtitle("Population vs Suicide Rate") +
geom_point(aes(x=population, y=suicides.100k.pop, color=year)) + xlab("Population") + ylab("Suicide Rate")
ggsave("population_suicide_rate.png")
```

``` {r}
#experiment with density plot 
ggplot(data)  + ggtitle("Density Plot of population vs suicide rate") +
geom_density_2d(aes(x=data$population, y=data$suicides.100k.pop, color=year)) + xlab("population") + ylab("suicide rate")
```

```{r}
var(data$population)
cor(data$suicides.100k.pop, data$population)
summary(data$population)
cor(data$population, data$depression_percentage)
cor(data$population, data$drug_death_rate)
```
```{r}
ggplot(data) + geom_line(aes(x = data$population, y = data$drug_death_rate)) + xlab("Population") +ggtitle("Scatterplot Drug Death Rate VS Population") + ylab("Drug Death Rate")
```
```{r}
ggplot(data) + geom_line(aes(x = data$population, y = data$depression_percentage)) + xlab("Population") +ggtitle("Scatterplot Depression Rate VS Population") + ylab("Depression Rate")
```
###Investigating Suicide rate and HDI for year

```{r}
var(data$HDI.for.year) #Very low variance for HDI year to year
cor(data$suicides.100k.pop, data$HDI.for.year) #Barley negatively correlated
cor(data$population, data$HDI.for.year)
cor(data$gdp_per_capita...., data$HDI.for.year)#Sanity check: has a positive correlation
cor(data$depression_percentage, data$HDI.for.year) 
cor(data$drug_death_rate, data$HDI.for.year) # correlation: 0.4429688 somewhat positivley correlated
summary(data$HDI.for.year)
as.numeric(names(table(data$HDI.for.year))[table(data$HDI.for.year) == max(table(data$HDI.for.year))]) # mode for HDI
```

```{r}
ggplot(data)  + ggtitle("HDI for year vs suicide rate") +
geom_jitter(aes(x=data$HDI.for.year, y=data$suicides.100k.pop)) + xlab("HDI yearly") + ylab("suicide rate")
```
###Investigating Suicide rate and GDP per capita

```{r}
ggplot(data)  + ggtitle("GDP per capita vs suicide rate") +
geom_line(aes(x=data$gdp_per_capita...., y=data$suicides.100k.pop, color = "gdp per capitia")) + 
xlab("GDP per capita") + ylab("suicide rate") + scale_color_manual(values=PALETTE[1:3])
```


```{r}
var(data$suicides.100k.pop, data$gdp_per_capita....)
cor(data$suicides.100k.pop, data$gdp_per_capita....)
summary(data$suicides.100k.pop, data$gdp_per_capita....)
```
```{r}
ggplot(data) + ylab("") + xlab("Suicide Rate")+ ggtitle("Lineplot, GDP and HDI vs Suicide Rate coded by color") +
      geom_line(aes(x=data$suicides.100k.pop, y= data$HDI.for.year, color = "HDI for year")) +
     geom_line(aes(x=data$suicides.100k.pop, y= data$gdp_per_capita...., color = "GDP Per Capita")) 

```

###Investigating Suicide rate and the drug death rate

```{r}
ggplot(data)+ geom_line(aes(x= data$drug_death_rate, y= data$suicides.100k.pop, color = "Drug Death Rate")) +
ylab("Suicide Rate") + xlab("Drug Death Rate")
```



```{r}
var(data$suicides.100k.pop, data$drug_death_rate)
cor(data$suicides.100k.pop, data$drug_death_rate)
cor(data$population, data$drug_death_rate)
cov(data$suicides.100k.pop, data$drug_death_rate)
summary(data$drug_death_rate)
```
###Investigating Suicide rate and the Depression Percentage

```{r}
ggplot(data)+ geom_line(aes(x= data$depression_percentage, y= data$suicides.100k.pop, color = "Depression Rate")) +
ylab("Suicide Rate") + xlab("Depression Percentage") + scale_color_manual(values=PALETTE[6])
```
Drug death rate and depression rate are slightly negitively correlated as well as suicide rates and drepression percentages.
```{r}
var(data$suicides.100k.pop, data$depression_percentage)
cor(data$suicides.100k.pop, data$depression_percentage)
cor(data$depression_percentage, data$drug_death_rate) 
cov(data$suicides.100k.pop, data$depression_percentage)
summary(data$depression_percentage)
```

```{r}
ggplot(data) + ylab("") + ggtitle("Lineplot, of Depression Rate, and Drug Death Rate v.s Suicide Rate") +
     geom_line(aes(x= data$suicides.100k.pop, y= data$depression_percentage, lty="Dashed", color = "Depression")) +
     geom_line(aes(x=data$suicides.100k.pop, y= data$drug_death_rate, lty="Solid", color = "Drug Death")) +
     #geom_line(aes(x=data$suicides.100k.pop, y= data$population, lty="x9", color = "Population")) +
     scale_linetype_manual(values=c("solid","longdash")) + xlab("Suicide Rate")
```

```{r}
temp_male <- data %>% filter(sex=="male")
temp_male$age <- factor(temp_male$age, levels = c("5-14 years", "15-24 years", "25-34 years", "35-54 years", "55-74 years", "75+ years"))
ggplot(data=temp_male) + geom_line(aes(x = year, y = suicides.100k.pop, group=age, color=age)) + xlab("Year") +ggtitle("Male Suicide Rates over Time by Age") + ylab("Suicide Rate")
ggsave("male_trend.png")
```

```{r}
temp_female <- data %>% filter(sex=="female")
temp_female$age <- factor(temp_female$age, levels = c("5-14 years", "15-24 years", "25-34 years", "35-54 years", "55-74 years", "75+ years"))
ggplot(data = temp_female) + geom_line(aes(x = year, y = suicides.100k.pop, group=age, color=age)) + xlab("Year") +ggtitle("Female Suicide Rates over Time by Age") + ylab("Suicide Rate")
ggsave("female_trend.png")
```
