---
title: "IEOR 142 FINAL PROJECT"
author: '3033342158'
date: "December 3, 2019"
output: html_document
---

#FINAL PROJECT IEOR 142
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(softImpute)
library(randomForest)
library(ranger)
library(dplyr)
library(reshape2)
library("caTools")
library(ROCR)
library(MASS)
library(ggplot2)
restore = list(repr.plot.width=8, repr.plot.height=3)
PALETTE = c("#00A9FF", "#F8766D", "#7CAE00", "#C77CFF", "#CD9600", "#00BE67", "#FF61CC", "#00BFC4")
theme.x_axis_only = theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.grid.major.y=element_blank(), panel.grid.minor.y=element_blank())
theme.no_legend = theme(legend.position="none")
theme.legend_title = theme(legend.title=element_text(size=7))
```



#EDA AND LOADING THE DATA
```{r}
data <- read.csv("kaggle.csv")
data.merged <- read.csv("suicide_rates.csv")
head(data)
tail(data)
```

There are 27820 observations in this dataset and 12 variables.
```{r}
nrow(data)
ncol(data)
```
The data contains 32 unique years from 1985 to 2016
```{r}
unique(sort(data$year))
length(unique(sort(data$year)))
```
There are 13910 males and 13910 females.
There is also 4642 occurances of every age range provided in the data except for the range 5-14 years which has only 4610.
```{r}
table(data$sex)
table(data$age) 
```
There are 744 observations correlating to the US and Japan, 372 per country.
```{r}
#Now let's explore the data corresponding to the US and Japan
nrow(data.merged)
head(data.merged)
#length(data.merged$country == "United States") 
#length(data.merged$country == "Japan")
nrow(data.merged[data.merged$country == "United States",])
nrow(data.merged[data.merged$country == "Japan",])
data.merged[data.merged$country == "United States",]
data.merged[data.merged$country == "Japan",]
```
This is not a very good visual here, since it is hard to interpret however it does seem clear that suicide has been steadily decsreasing across several countries as years go on.
```{r}
ggplot(data.merged) + ggtitle("Suicide Rate Per Year") + geom_line(aes(x= data.merged$year, y=data.merged$suicides.100k.pop)) +
xlab("Year") + ylab("suicide #")

```

```{r}
ggplot(data.merged) + ggtitle("Suicide Number Per age range") + geom_line(aes(x= data.merged$age, y=data.merged$suicides.100k.pop)) +
xlab("Age Range") + ylab("suicide rate")


```
The age range that most suicides seem to occur is 75+ years of age.
```{r}
ggplot(data.merged) + ggtitle("Suicide Number Per age range") + geom_boxplot(aes(x= data.merged$age, y=data.merged$suicides.100k.pop)) +
xlab("Age Range") + ylab("suicide #")
ggplot(data.merged) + ggtitle("Suicide Number Per age range") + geom_col(aes(x= data.merged$age, y=data.merged$suicides.100k.pop)) +
xlab("Age Range") + ylab("suicide rate")

```
The smaller the population size the higher the suicide number.
```{r}
ggplot(data.merged)  + ggtitle("Population vs Suicide Rate (per 100K)") +
geom_point(aes(x=data.merged$population, y=data.merged$suicides.100k.pop, color=country)) + xlab("population") + ylab("suicide rate")

ggplot(data.merged)  + ggtitle("population vs suicide rate") +
geom_area(aes(x=data.merged$population, y=data.merged$suicides.100k.pop)) + xlab("population") + ylab("suicide rate")

```

```{r}
copy <- data.frame(data.merged)
copy$age <- factor(copy$age, levels = c("5-14 years", "15-24 years","25-34 years","35-54 years","55-74 years","75+ years"))

ggplot(copy)  + ggtitle("Suicide Rate by Year") +
geom_line(aes(x=data.merged$year, y=data.merged$suicides.100k.pop, color=country)) + xlab("year") + ylab("suicide rate") + facet_wrap(vars(age, sex))
aspect_ratio <- 2.5
ggsave("suicide_rate_eda.png", height = 7 , width = 5 * aspect_ratio)

```
There seems to be a negative correlation of  0.5216032 between the suicide rate and population and suicide rate and gdp per capita.
```{r}
cor(data.merged$suicides.100k.pop, data.merged$population)
cor(data.merged$suicides.100k.pop, data.merged$gdp_per_capita....)
```

The Russian Federation has the highest recorded suicide number in the years that our data has, with 1994 being the year of occurance.
The country with the lowest record is Albania(the data may be off for this since it is 0), Cyprus, Grenada and Antigua and Barbuda, and Iceland seem to all have very low suicude numbers.

However for suicide rate in preportion to population size the highest recorded suicide rate occured in Aruba in 1995 (a 75+ male). Min is still Albania.


```{r}
max(data$suicides.100k.pop)
min(data$suicides.100k.pop)
data[data$suicides.100k.pop == max(data$suicides.100k.pop), ]
data[data$suicides.100k.pop == min(data$suicides.100k.pop), ]

data[data$suicides.100k.pop == min(data$suicides.100k.pop) & data$year == "2016" , ] # Countires that have the lowest suicide numbers in recent years(2016) also checked 2015
data[data$suicides.100k.pop == min(data$suicides.100k.pop) & data$year > "2013" , ]

```
The below visualization shows suicide is much more common in males than females in the US.
```{r}
us.data <- data.merged[data.merged$country == "United States",]

ggplot(us.data)  + ggtitle("suicide number per age for the US") +
geom_col(aes(x=us.data$sex, y=us.data$suicides.100k.pop)) + xlab("Sex") + ylab("suicide number") 
```

The data shows there was a decrease in the suicide rate from around the years 1999 to 2010 but since 2010 has been a steady increasing in the US.
```{r}
ggplot(us.data)  + ggtitle("suicide trend over the years for the US") +
geom_col(aes(x=us.data$year, y=us.data$suicides.100k.pop)) + xlab("Year") + ylab("suicide number") 

```
The generation with the highest suicide rate is Silent generation in the US.

```{r}
ggplot(us.data)  + ggtitle("suicide trend over generations for the US") +
geom_col(aes(x= us.data$generation, y= us.data$suicides.100k.pop)) + xlab("generation") + ylab("suicide number")
```


```{r}
Japan.data <-  data.merged[data.merged$country == "Japan",]
```
For the most part, the data shows there has been a steady decrease in the number of suicides in the Japan there was also a was a large decline in suicide rate for most of the 1990's time period.

```{r}
ggplot(Japan.data)  + ggtitle("suicide trend over the years for the Japan") +
geom_col(aes(x= Japan.data$year, y= Japan.data$suicides.100k.pop)) + xlab("Year") + ylab("suicide rate") 

```
Suicide among males is also more common in Japan than in females.
```{r}
ggplot(Japan.data)  + ggtitle("suicide number per age for the Japan") +
geom_col(aes(x=Japan.data$sex, y=Japan.data$suicides.100k.pop)) + xlab("Sex") + ylab("suicide number")
```
Similarly to the US, Japan's suicide rate is highest among the Silent generation.
```{r}
ggplot(Japan.data)  + ggtitle("suicide trend over generations for the Japan") +
geom_col(aes(x= Japan.data$generation, y= Japan.data$suicides.100k.pop)) + xlab("generation") + ylab("suicide number")
```
Japan has a highest recorded suicide rate out of Japan and the US given the years, occuring in 1986. 
```{r}
data.merged[data.merged$suicides.100k.pop == min(data.merged$suicides.100k.pop), ]
data.merged[data.merged$suicides.100k.pop == max(data.merged$suicides.100k.pop), ]

```



```{r}
#plot(data.merged[data.merged$suicides_no == max(data.merged$suicides_no), ])

```






