---
title: "timeseries"
author: '3032437864'
date: "12/13/2019"
output: html_document
---

# Time series
```{r}
# R^2 with a particular baseline
BaselineR2 <- function(predictions, truth, baseline) {
  SSE <- sum((truth - predictions)^2)
  SST <- sum((truth - baseline)^2)
  r2 <- 1 - SSE/SST
  return(r2)
}
```

```{r}
# Use 2013 as testing data
train_hourly <- us %>% filter(Year < 2013)
test_hourly <- us %>% filter(Year >= 2013)
train_hourly
```

```{r}
set.seed(377)

# Plot initial data -- ggplot knows how to plot dates!
melted = reshape2::melt(us %>% select(c(Year, Total..16.years.and.older, White, Black.or.African.American, Asian, Hispanic.or.Latino.ethnicity)), id.var='Year')
ggplot(melted, aes(x=Year, y=value, color=variable)) + geom_point() + scale_y_continuous(breaks = round(seq(min(melted$value), max(melted$value), by = 2),1))
```

####BUILDING MODELS FOR INDIVIDUAL COLUMNS: FIRST COL IS WHITE
```{r}

# Linear trend model training data -- Make a new column for the time period
# number (1, 2, ...). The dplyr syntax is a little tricky here -- n() is the
# number of rows in salesTrain, and seq_len(n()) returns the vector 1, 2, ...,
# n(). The end result is that we added a new variable called TimePeriod that
# takes values 1, 2, ..., n().
trainLM_hourly <- train_hourly %>% mutate(TimePeriod = seq_len(n()))
# Build and plot linear trend model
modLM <- lm(White~TimePeriod, data=trainLM_hourly)
ggplot(trainLM_hourly, aes(x=Year, y=White)) +
  geom_line() +
  geom_point() +
  geom_line(aes(y=predict(modLM)), col="red", lwd=1.5)
```

###### Random Walk model training data 
```{r}
trainRW_hourly <- train_hourly %>% mutate(LastYear = c(NA, head(White, -1)))
head(trainRW_hourly)
#random walk aka moving average
```

```{r}
# Plot with an additional red line for our predictions as before
ggplot(trainRW_hourly, aes(x=Year, y=White)) +
  geom_line() +
  geom_point() +
  geom_line(aes(y=LastYear), col="red")
```

```{r}
# Proportion of percentages for which difference is more than 1.
table(abs(trainRW_hourly$White-trainRW_hourly$LastYear) >= 1)
# Compute training set R^2
# Note that we need to remove the first observation since there is no
# prediction. This is achieved using tail(.., -1) which says to take all but
# the first observation.
BaselineR2(tail(trainRW_hourly$LastYear, -1), 
           tail(trainRW_hourly$White, -1),
           mean(trainRW_hourly$White))
```


###### AR model
```{r}
# We need to add sales yesterday and sales two days ago for the two term AR model
# head(.., -2) says take all but the last two 
trainAR_hourly <- train_hourly %>%
  mutate(LastYear=c(NA, head(White, -1))) %>%
  mutate(TwoYearsAgo = c(NA, NA, head(White, -2)))
# Do the regression with one lag term
mod2a <- lm(White~LastYear, data=trainAR_hourly)
summary(mod2a)
```

# 2-term autoregressive model 
mod2b <- lm(White~LastYear+TwoYearsAgo, data=trainAR_hourly)
summary(mod2b)
# Plot with an additional red line for our predictions as before
ggplot(trainAR_hourly, aes(x=Year, y=White)) +
  geom_line() +
  geom_point() +
  geom_line(aes(y=predict(mod2b, newdata=trainAR_hourly)), col="red")


## Trying Random Forest
library(randomForest)
set.seed(349)

# Plug in all of the variables that we've created
mod.rf <- randomForest(White ~ LastYear + TwoYearsAgo + Year, data = tail(trainAR_hourly, -2))
ggplot(trainAR_hourly, aes(x=Year, y=White)) +
  geom_line() +
  geom_point() +
  geom_line(aes(y=predict(mod.rf, newdata=trainAR_hourly)), col="green")

# Both on the same plot:
ggplot(trainAR_hourly, aes(x=Year, y=White)) +
  geom_line() +
  geom_point() +
  geom_line(aes(y=predict(mod2b, newdata=trainAR_hourly)), col="red") +
  geom_line(aes(y=predict(mod.rf, newdata=trainAR_hourly)), col="green")

# Create Test Set
test_hourly_final <- test_hourly %>%
  mutate(LastYear=c(NA, head(White, -1))) %>%
  mutate(TwoYearsAgo = c(NA, NA, head(White, -2)))

# Test set prediction and OSR^2
# Test-set prediction
pred.test <- tail(predict(mod2b, newdata = test_hourly_final), -2)
OSR2(pred.test, trainAR_hourly$White, tail(test_hourly_final$White, -2))

pred.test.rf <- tail(predict(mod.rf, newdata = test_hourly_final), -2)
OSR2(pred.test.rf, trainAR_hourly$White, tail(test_hourly_final$White, -2))

##mod2b does slightly better in terms of OSR^2, but that might be because of limited data
# we should test with a greater fraction in test set or go with random forest maybe?

# Test set plots
ggplot(test_hourly_final, aes(x=Year, y=White)) +
  geom_line() +
  geom_point() +
  geom_line(aes(y=pred.test), col="red")

ggplot(test_hourly_final, aes(x=Year, y=White)) +
  geom_line() +
  geom_point() +
  geom_line(aes(y=pred.test), col="red") +
  geom_line(aes(y=pred.test.rf), col="green")
```

